---
title: "Dirty Cafe Sales"
output: html_document
date: "2025-03-31"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This report dives into a synthetic dataset about Cafe Sales. This dataset was retrive from Kaggle and was uploaded by Ahmed Mohamed in January. The data includes transaction history of purchased items at the cafe along with the cafe items. This contains 10,000 rows of synthetic data that needs to be cleaned. There are issues, such as incorrect data types, missing values, and inconsistent totals. The goal is to clean the data for accurate analysis.

You can visit [Kaggle](https://www.kaggle.com/datasets/ahmedmohamed2003/cafe-sales-dirty-data-for-cleaning-training/data) for more information about the data.

```{r, include = FALSE}
library(tidyverse)
library(dplyr)
```

## Load Data

Let's load the dataset and review the data.

```{r}
#Read in the dataset
dirty_sales <- read_csv("dirty_cafe_sales.csv")

#View the dataset
glimpse(dirty_sales)
```

```{r, include=FALSE}
```

## Data Cleaning

We will perform the following steps to clean the data:

1.  Convert incorrect data types
2.  Handle missing values
3.  Ensure total spent is calculated correctly

### Convert Data Types

When reviewing the data, you may have noticed that some numerical columns were stored as characters. Lets fix this!

To do this, we need to  fix the incositent format in some columns that include "ERROR", "UNKOWN", and "NA". We will change it all to "NA" for consistency.

```{r}
clean_sales <- dirty_sales |>  
  mutate(
    `Price Per Unit` = na_if(`Price Per Unit`, "ERROR|UNKOWN"),
    `Total Spent` = na_if(`Total Spent`, "ERROR|UNKOWN"),
    Quantity = na_if(Quantity, "ERROR|UNKOWN"),
    Item = na_if(Item, "ERROR|UNKNOWN")
  )
```

Now that this is resolved, we can change the columns to a different data type.

```{r, warning=FALSE}
clean_sales <- dirty_sales |>
  mutate(
    `Price Per Unit` = as.double(`Price Per Unit`),
    `Total Spent` = as.double(`Total Spent`),
    Quantity = as.integer(Quantity),
    `Transaction Date` = as.Date(`Transaction Date`)
    )

glimpse(clean_sales)
```

```{r, include=FALSE}

```

### Handle Missing Values

We now need to identify and handle missing values. We will first handle the numerical missing values. 

```{r}
clean_sales <- clean_sales |>
  mutate(
    `Total Spent` = ifelse(is.na(`Total Spent`), Quantity * `Price Per Unit`, `Total Spent`),
    `Price Per Unit` = ifelse(is.na(`Price Per Unit`),`Total Spent` / Quantity, `Price Per Unit`),
    Quantity = ifelse(is.na(Quantity),`Total Spent` / `Price Per Unit`, Quantity)
  )
```

Now lets look into the item missing values. I created a price reference chart to make sure that all items in the clean_sales dataframe are correct.

```{r, warning=FALSE}
#Handle Item missing values
price_reference <- tibble(
  Item = c("Coffee", "Tea", "Sandwich", "Salad", "Cake", "Cookie", "Smoothie", "Juice"),
  Price = c(2.0, 1.5, 4.0, 5.0, 3.0, 1.0, 4.0, 3.0 )
)

#This makes sure all items prices are correct
clean_sales <- left_join(clean_sales, price_reference, by = c("Price Per Unit" = "Price")) |>
  relocate(Item.y, .before = "Quantity") |>
  #select(c(1:8) ) |> # need to delete two columns
  rename("Item" = "Item.y") |>
  distinct()

glimpse(clean_sales)
```

In the Payment Method and Location columns you may have noticed there are "UNKOWN" and "ERROR" values. Since we do not have additional information to clean this, we will remove the rows with this data.
```{r, warning=FALSE}
clean_sales <- clean_sales |>
  filter(!is.na(`Payment Method`), `Payment Method` != "UNKNOWN", `Payment Method` != "ERROR") |>
  filter(!is.na(Location), Location != "ERROR") 

clean_sales <- clean_sales |>
  arrange(`Transaction Date`)
clean_sales_quantity <- clean_sales |>
  filter(!is.na(Quantity), !is.na(`Total Spent`), !is.na(`Price Per Unit`)) |>
  group_by(Item) |>
  summarise(Item_total = sum(`Total Spent`))

glimpse(clean_sales)

```

```{r, include=FALSE}

```

With our clean data, let's dive deeper into it. Lets look at the total sales each cafe item brings in. This is shown below.

```{r}
ggplot(clean_sales_quantity, aes(x = Item, y = Item_total)) +
  geom_bar(stat = "identity", fill = "darkgreen") +
  labs(
    title = "Total Sales By Cafe Item",
    x = "Item",
    y = "Total Sales ($)"
  )
```

The bar chart visualizes the total revenue made by each cafe item. This helps identify which item is the most profitable and which item may need sales improvement. From the chart, we can see that, sandwiches and smoothies have the top revenue generating items. This can be due to the constant demand of these items or the price as well. On the other hand, cookies have low sales. This can be due to the lack of intrest from customers or a myraid of other reasons. 

This data provides valuable insights for the cafe's next steps. Top selling items should be prioritized in restocking and selling strategies. Cookies might benefit from having promotional deals or being removed from the menu. 

## Conclusion

We have successfully cleaned the Dirty Cafe Sales dataset by:

1. Converting data types
2. Handling missing values
3. Fixing total spent inconsistencies
4. Standardizing categorical values

This cleaned dataset is now ready for further analysis.
